### 2013-11-10 src2/thesis.tex
spell pass

I would like to thank my supervisor Mary Sheeran. She is a very suportive, motivating and 
understanding supervisor. I also thank my co-supervisor Koen Claessen for all the great advice 
and tips he has provided over the years. I cannot imagine that this would have worked with 
any other supervisors. Thanks to Josef Svenningsson, a late addition to my troop of 
supervisors. Working with Josef has been fun and rewarding.

@@

I want to thank Erik, Ola, Markus and Viktor for our occasional lunches and GPU discussions.

@@

@@

Obsidian is geared towards the programmer who is interested in GPU hardware. Making 
good use of Obsidian requires that the programmer is willing to make an effort 
to learn about GPUs and how to program them efficiently. However, another case 
is a programmer who does not want to invest that effort but wants to write a high-level 
parallel program and have it run reasonably well on different parallel architectures. 
In this case, a language with even higher level of abstraction is preferable. It then 
becomes the job of a more advanced compiler to turn that into code for the targetted 
platforms. The EmbArBB language targets this programmer. EmbArBB exposes a set of high-level 
operations on arrays, such as {\tt map}, {\tt reduce} and {\tt scan}. The programmer 
has no power over the parallel strategy that should be used to implement these 
operations but instead relies on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
Haskell embedding of Intel ArBB that was released as an embedded language in C++. 
EmbArBB relies entirely on the ArBB system for code generation and optimisation.

@@

\noindent The kernel takes two input arrays, {\tt i1} and {\tt i2}; the result is stored 
in {\tt r}. Note how the kernel makes use of three variables, {\tt threadIdx.x}, 
{\tt blockIdx.x} and {\tt gridDim.x}. These variables identify a thread uniquely within 
a grid. Here they are used to obtain an index into the provided arrays. The dimensions of 
the blocks and of the grid are set by a controling program upon launch of the kernel. This 
controling program is run by the CPU of the computer housing the GPU. The kernel is launched 
using the following syntax:

@@

The function above takes a pull array and creates a push array with a parallel 
iteration schema over threads within a block. There are other iteration schemas to 
chose from such as sequential, combinbations of sequential and parallel
across both threads and blocks (two level parallelism).

Push arrays give many of the same benefits that pull arrays have. Using push 
arrays we also get fusion of operations as default. The implementation of map 
(shown below) illustrates how maping a function {\tt f} over a push array results 
in an application of {\tt f} in the write-function.

@@

In paper E (section~\ref{sec:paperE}), we implement counting sort and occurrance sort 
using Obsidian. We use very low-level features of Obsidian and end up producing 
code that performs very well.

@@

Meta-Repa~\citet{METAREPA} is a reimplementation of the Repa library using deep embedded language 
techniques. Meta-Repa programs build ASTs that are then compiled (using template 
Haskell) to Haskell code using low-level parts of the Repa library for its parallelism.
The Meta-Repa approach to implementing the Repa library 
gives inlining of operations for free, where Repa programs need to be annotated 
with pragmas to ensure inlining. On the other hand, in Meta-Repa the programmer 
needs to explicitly state when something should not be inlined (using a {\tt force} primitive).
This is related to our work on EmbArBB that also uses an existing library (ArBB) 
for its parallelism. In reference~\citet{FPCDSL}, a case is made for this style 
of embedding prexisting DSLs.

@@

``Compiling Embedded Languages'' by Elliott and Finne has been a constant source of information 
and something we have refered to in almost all our work. This is, in my opinion, the paper to turn 
to for anyone starting out with compiled embedded languages.

@@

Obsidian is a language for GPU kernel implementation, not complete applications. 
It is not as well integrated into Haskell as for example Accelerate. There are 
ways to execute Obsidian kernels from within Haskell (we have seen examples of this in 
section~\ref{sec:GPUPROGRAMMING}); the recommended workflow, 
however, is to generate the CUDA as text and write the host code by hand. This 
can be seen as both a shortcoming and a feature. It is hard to integrate 
Accelerate programs in larger applications written in other languages; when 
using Obsidian a kernel is generated and can be used as part of for example a library.

@@

This thesis contains two approaches to embedding a language for data-parallel programming 
in Haskell. EmbArBB embeds a preexisting library for data-parallelism in Haskell 
and at little effort from us, very good performance is gained. The performance of 
EmbArBB is entirely due to all the optimisation expertise and effort that was invested in 
the ArBB system. We get all that expertice for free. What EmbArBB provides is a 
way for Haskellers to benefit from the ArBB system. ArBB and EmbArBB are based on a 
set of parallel primitives for the programmer to use and compose. Obsidian takes 
a different approach. Built in primitives are very few and very low level, such as 
sequential and parallel for loops. On top of the low-level capabilities, a library 
of high-level primitives can be built. The key is that we want the programmer to 
be able to specify how to implement these higher level abstractions.

>>>>

I would like to thank my supervisor Mary Sheeran. She is a very supportive, motivating and 
understanding supervisor. I also thank my co-supervisor Koen Claessen for all the great advice 
and tips he has provided over the years. I cannot imagine that this would have worked with 
any other supervisors. Thanks to Josef Svenningsson, a late addition to my troop of 
supervisors. Working with Josef has been fun and rewarding.

I have often turned to Emil Axelsson with technical questions and he has always been
happy to help. For this I am grateful.

@@

@@

I want to thank Erik, Ola, Markus and Viktor for our occasional lunches and GPU discussions. 
I also thank my office mates, Michal and Nicholas, for their company. I thank Eva Axelsson, 
who I have felt most comfortable to turn to for help and information, even when she was not 
the appropriate channel.

@@

Obsidian is geared towards the programmer who is interested in GPU hardware. Making 
good use of Obsidian requires that the programmer is willing to make an effort 
to learn about GPUs and how to program them efficiently. However, another case 
is a programmer who does not want to invest that effort but wants to write a high-level 
parallel program and have it run reasonably well on different parallel architectures. 
In this case, a language with even higher level of abstraction is preferable. It then 
becomes the job of a more advanced compiler to turn that into code for the targeted 
platforms. The EmbArBB language targets this programmer. EmbArBB exposes a set of high-level 
operations on arrays, such as {\tt map}, {\tt reduce} and {\tt scan}. The programmer 
has no power over the parallel strategy that should be used to implement these 
operations but instead relies on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
Haskell embedding of Intel ArBB that was released as an embedded language in C++. 
EmbArBB relies entirely on the ArBB system for code generation and optimisation.

@@

\noindent The kernel takes two input arrays, {\tt i1} and {\tt i2}; the result is stored 
in {\tt r}. Note how the kernel makes use of three variables, {\tt threadIdx.x}, 
{\tt blockIdx.x} and {\tt gridDim.x}. These variables identify a thread uniquely within 
a grid. Here they are used to obtain an index into the provided arrays. The dimensions of 
the blocks and of the grid are set by a controlling program upon launch of the kernel. This 
controlling program is run by the CPU of the computer housing the GPU. The kernel is launched 
using the following syntax:

@@

The function above takes a pull array and creates a push array with a parallel 
iteration schema over threads within a block. There are other iteration schemas to 
chose from such as sequential, combinations of sequential and parallel
across both threads and blocks (two level parallelism).

Push arrays give many of the same benefits that pull arrays have. Using push 
arrays we also get fusion of operations as default. The implementation of map 
(shown below) illustrates how mapping a function {\tt f} over a push array results 
in an application of {\tt f} in the write-function.

@@

In paper E (section~\ref{sec:paperE}), we implement counting sort and occurrence sort 
using Obsidian. We use very low-level features of Obsidian and end up producing 
code that performs very well.

@@

Meta-Repa~\citet{METAREPA} is a reimplementation of the Repa library using deep embedded language 
techniques. Meta-Repa programs build ASTs that are then compiled (using template 
Haskell) to Haskell code using low-level parts of the Repa library for its parallelism.
The Meta-Repa approach to implementing the Repa library 
gives inlining of operations for free, where Repa programs need to be annotated 
with pragmas to ensure inlining. On the other hand, in Meta-Repa the programmer 
needs to explicitly state when something should not be inlined (using a {\tt force} primitive).
This is related to our work on EmbArBB that also uses an existing library (ArBB) 
for its parallelism. In reference~\citet{FPCDSL}, a case is made for this style 
of embedding preexisting DSLs.

@@

``Compiling Embedded Languages'' by Elliott and Finne has been a constant source of information 
and something we have referred to in almost all our work. This is, in my opinion, the paper to turn 
to for anyone starting out with compiled embedded languages.

@@

Obsidian is a language for GPU kernel implementation, not complete applications. 
It is not as well integrated into Haskell as for example Accelerate. There are 
ways to execute Obsidian kernels from within Haskell (we have seen examples of this in 
section~\ref{sec:GPUPROGRAMMING}); the recommended work flow, 
however, is to generate the CUDA as text and write the host code by hand. This 
can be seen as both a shortcoming and a feature. It is hard to integrate 
Accelerate programs in larger applications written in other languages; when 
using Obsidian a kernel is generated and can be used as part of for example a library.

@@

This thesis contains two approaches to embedding a language for data-parallel programming 
in Haskell. EmbArBB embeds a preexisting library for data-parallelism in Haskell 
and at little effort from us, very good performance is gained. The performance of 
EmbArBB is entirely due to all the optimisation expertise and effort that was invested in 
the ArBB system. We get all that expertise for free. What EmbArBB provides is a 
way for Haskellers to benefit from the ArBB system. ArBB and EmbArBB are based on a 
set of parallel primitives for the programmer to use and compose. Obsidian takes 
a different approach. Built in primitives are very few and very low level, such as 
sequential and parallel for loops. On top of the low-level capabilities, a library 
of high-level primitives can be built. The key is that we want the programmer to 
be able to specify how to implement these higher level abstractions.
